---
title: "Basic Use of `lottery`"
author: "Pierre DENELLE"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Use of `lottery`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Aim of `lottery`

Quick and efficient simulation of communities undergoing both neutral and niche-based assembly processes. The coalescent-based approach provides local community composition based on the genealogy of its individuals. It does not require simulating community dynamics from an initial state forward in time, and the genealogical approach can incorporate the influence of environmental filtering and species niche differences. The sampling process allows simulating varying sample size in a flexible way.

# Usage

## Regional pool generation

We can use the function coalesc to construct a regional $pool$ of species. Otherwise, the user can provide a regional $pool$ with individuals, species and trait values of species (optional).

By default, species abundances will follow a log-serie distribution and if no trait values will be sampled from uniform distribution.

```{r}

# Generate a regional pool/metacommunity with log-series distribution of abundances and uniform trait distribution
pool <- coalesc(J = 25000, m = 1, theta = 50)$pool

par(mfrow = c(1, 1))
hist(table(pool$sp), xlab = "Species", main = "Distribution of species' abundances")
hist(pool$tra1, xlab = "Trait values", main = "Distribution of trait values")

```

## Community from a given pool

$coalesc$ function constructs a community from any regional pool. In this example, we provide a regional pool with a uniform regional abundances distribution. Species also have a trait value.
Here we simulate a set of 10 communities from this same regional pool.

We get a data frame with each individual occurring in the focal community, its species label and its trait value.
To get the abundances of species, we can use the $abund$ function of the package.

```{r, include=FALSE}

#Generate a regional pool/metacommunity with equal abundances and uniform trait distribution
pool <- cbind(1:100000, rep(sample(1:100), 1000), rep(NA, 100000)) # 100 species
t.sp <- runif(100)
pool[,3] <- t.sp[pool[,2]]

com_df <- c()

for (i in 1:10){
  comm <- coalesc(J = 200, m = 0.9, theta = 50, pool = pool, traits = NULL)
  com_df <- rbind(com_df, cbind(rep(i, nrow(comm$com)), comm$com))
}

colnames(com_df) <- c("com", "ind", "sp", "tra1")

```

```{r}

head(abund(comm)$com)

```

## Use of environmental filters

When drawing species from regional pool, an environmental filtering can arise and determine the probability of a species to occur in a community according its trait value.

Here we show two examples where two filters happen. Both select the species according the proximity of their trait value to a local optimum called topt. The first one is stabilizing since it selects favourably the species with a trait value close to the local topt. The second one is disruptive and produces the opposite pattern.

```{r}

filt_disruptive <- function(t, x, sigma) 1-exp(-(x-t)^2/(2*sigma^2))
filt_gaussian <- function(t, x, sigma) exp(-(x-t)^2/(2*sigma^2))
filt <- function(x) filt_gaussian(0.5, x[1], sigma = 0.1) * filt_disruptive(0.5, x[2], sigma = 0.1)

# Example 1
# Parameters of coalesc
J <- 500
Jpool <- 50*J
m <- 1
theta <- 50
# A regional pool including 500 species each including 50 individuals
pool <- cbind(1:Jpool, rep(1:500, 50))
# 2 traits distributed uniformly
traits <- cbind(runif(Jpool), runif(Jpool)) 
# Simulating one community
ex <- coalesc(J, m, theta, filt = filt, traits = traits[, 1:2], pool = pool)
# Plots
par(mfrow=c(1, 3))
plot(tapply(ex$com$tra1, ex$com$sp,length) ~ tapply(apply(ex$com[, c("tra1", "tra2")], 1, filt), ex$com$sp, mean),
     xlab = "Filtering", ylab = "Abundance")
plot(tapply(ex$com$tra1, ex$com$sp,length) ~ tapply(ex$com$tra1, ex$com$sp, mean),
     xlab = "Value of first trait", ylab = "Abundance")
plot(tapply(ex$com$tra1, ex$com$sp,length) ~ tapply(ex$com$tra2, ex$com$sp, mean),
     xlab = "Value of second trait", ylab = "Abundance")


# Example 2
# Parameters of coalesc
m <- 1
theta <- 50
# Generate a regional pool/metacommunity with log-series distribution of abundances and uniform trait distribution
pool <- coalesc(J = 25000, m = m, theta = theta)$pool
# Add a second trait in regional pool, partially correlated with the first one
tracor <- tcor(n = 25000, rho = 0.8, x = pool$tra1, mar.fun = runif)
par(mfrow = c(1, 1))
plot(tracor[, 1], tracor[, 2])
cor(tracor[, 1], tracor[, 2])
pool$tra2 <- tracor$t2
# Simulating one community
ex <- coalesc(J, m, theta, filt = filt, traits = pool[, c("tra1", "tra2")], pool = pool[, c("ind", "sp")])
# Plots
par(mfrow=c(1, 3))
plot(tapply(ex$com$tra1, ex$com$sp,length) ~ tapply(apply(ex$com[, c("tra1", "tra2")], 1, filt), ex$com$sp, mean),
     xlab = "Filtering", ylab = "Abundance")
plot(tapply(ex$com$tra1, ex$com$sp,length) ~ tapply(ex$com$tra1, ex$com$sp, mean),
     xlab = "Value of first trait", ylab = "Abundance")
plot(tapply(ex$com$tra1, ex$com$sp,length) ~ tapply(ex$com$tra2, ex$com$sp, mean),
     xlab = "Value of second trait", ylab = "Abundance")

```

# Community visualizations

$lottery$ contains a preset function called $plot_comm$ to visualize the distribution of trait and abundances in the regional pool and one community drawn from this regional pool.

```{r}

plot_comm(ex)
par(mfrow = c(1, 1))
plot_comm(ex, type = "abund")

```

# Comparison with forward approach

A second way to construct communities is implemented in $lottery$ package. $forward$ function starts from an initial state and then draws death, birth and arrival by dispersal of individuals over a given number of successive generations. 

Several parameters can be implemented, including environmental filtering as described above for the coalesc function.

$forward$ function provides a list with two elements:
- a $data.frame$ called $com$ describing the final community, with individuals and their species labels and trait values
- a $data.frame$ called $pool$ describing the regional pool of individuals used to draw the communities

If $keep$ parameter is set to $TRUE$, the object contains three elements:
- a $list$ called $com_t$ which contains all the communities generated per iteration step
- a $vector$ called $limit.sim.t$ containing the average value of limiting similarity over each generation
- a $data.frame$ called $pool$ describing the regional pool of individuals used to draw the communities

Trait distributions can be viewed with $plot_comm$ function.

```{r}

# Disruptive environmental filtering around topt = 0.5
filt_disruptive <- function(t, x, sigma) 1-exp(-(x-t)^2/(2*sigma^2))
filt <- function(x) filt_disruptive(0.5, x, sigma = 0.1)

# A regional pool including 100 species each including 10 individuals
pool <- sort(rep(as.character(1:100), 10))

# Initial community composed of 10 species each including 10 individuals, 
# with trait information for niche-based dynamics
initial <- data.frame(sp = sort(rep(as.character(1:10), 10)), trait = runif(100))

# Variation of community richness with time
final <- forward(initial = initial2, prob = 0.1, gens = 1000, pool = pool, 
                 limit.sim = TRUE, keep = TRUE, filt = filt)

final2 <- forward(initial = initial2, prob = 0.1, gens = 1000, pool = pool, 
                 limit.sim = TRUE, keep = F, filt = filt)

# Check stationarity
plot(unlist(lapply(final$com_t, function(x) length(unique(x[, 2])))), 
     xlab = "Time step", ylab = "Community richness") 

# Index of limiting similarity over time
plot(final$limit.sim.t, xlab = "Time step", ylab = "Limiting similarity")

# Comparison of regional and local trait distributions
plot_comm(list("pool" = final$pool, "com" = final$com_t[1000][[1]]))
plot_comm(final2)

```

The use of $forward$ function then produces the same pattern than $coalesc$ function but differs in two aspects:
- it allows the user to see in details the processes happening at each generation
- it is more time and memory consuming


